---
title: "Conducting Power Analysis for Haddock Epigenetic Aging Project"
author: "Authors: Emma Strand; emma.strand@gmgi.org"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

# Conducting a power analysis to decide sample size 

Information can also be found here: https://github.com/emmastrand/GMGI_Notebook/blob/main/posts/2023-08-09_PowerAnalysis.md. 

Tutorials followed: 

- https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html   
- https://cran.r-project.org/web/packages/pwr/vignettes/pwr-vignette.html  

## Install and load pwrss R package 

Bulus, M. (2023). pwrss: Statistical Power and Sample Size Calculation Tools. R package version 0.3.1. https://CRAN.R-project.org/package=pwrss

Bulus, M., & Polat, C. (in press). pwrss R paketi ile istatistiksel guc analizi [Statistical power analysis with pwrss R package]. Ahi Evran Universitesi Kirsehir Egitim Fakultesi Dergisi. https://osf.io/ua5fc/download/

```{r}
#install.packages("pwrss") ##commented once I downloaded this so it won't run every time 
#install.packages("pwr") ##commented once I downloaded this so it won't run every time 
library(pwrss)
library(pwr)
```

I'm looking at two types of analysis here: Correlation and Linear Regressions. Both quantify direction and strength of the relationship between two numeric variables (otolith age and epigenetic age).   
- Correlation = how well the variables are linearly related. Has value -1 to 1.    
- Linear regression = estimates parameters of a linear equation that can be used to predict one variable based on another. Can address cause and effect, use equation to predict value of one variable based on another, equation to quantify relationship.



## Linear Regression 

https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html#3_Linear_Regression_(F_and_t_Tests)  
https://cran.r-project.org/web/packages/pwr/vignettes/pwr-vignette.html 

Tests model fit: 
- Omnibus F test in multiple liner regression is used to test whether R2 is greater than 0 (zero).   
- Hierarchial Linear Regression tests  multiple factors within a regression analysis.  

We are expecting that these three variables explain 90% of the variance in the outcome (R2=0.90 or r2 = 0.90 in the code). Based on this, what is the minimum required sample size?

- `alpha` = significance level
- `r2` = R2 for a linear regression  
- `k` = number of variables
- `power` = power of the test (0.8 is minimum to be considered high power)

### Omnibus F test

https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html#3_Linear_Regression_(F_and_t_Tests)  

```{r}
pwrss.f.reg(r2 = 0.90, k = 2, power = 0.80, alpha = 0.001)
```

### Hierarchial Linear Regression 

https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html#3_Linear_Regression_(F_and_t_Tests)  

- `m` = # of factors added 

```{r}
pwrss.f.reg(r2 = 0.90, k = 5, m = 2, power = 0.80, alpha = 0.05)
```
### General Linear Model

https://cran.r-project.org/web/packages/pwr/vignettes/pwr-vignette.html

- `f2` = effect size where f2 = R2 / (1-R2)  
- `u` = numerator degrees of freedom; number of coefficients you'll have in your model (minus the intercept)

Sample size (n) = v + u + 1
From below command = 5.948 + 2 + 1 = 8.948 ~9 samples with power = 0.8
From below command = 6.592 + 2 + 1 = 9.5292 ~10 samples with power = 0.9

```{r}
pwr.f2.test(u = 2, f2 = 0.9/(1 - 0.9), sig.level = 0.001, power = 0.9)
```

## Correlation Test 

https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html#10_Correlation(s)_(z_Test)

- `r` = expected correlation
- `r0` = value expected to be different from r  
- `alternative` = "not equal" indicates a two-sided test and "greater" indicates a one-sided test
- `alpha` = significance value

```{r}
pwrss.z.corr(r = 0.2, r0 = 0,
             power = 0.90, alpha = 0.0001, 
             alternative = "not equal")
```

## Ending notes 

With significance level < 0.001 and power = 0.9, we need at least ten samples in a linear regression and correlation. We'll have way more than this so this won't be a worry. 





