---
title: "Sea Urchin CpG O/E Calculations"
author: "Author: Emma Strand; emma.strand@gmgi.org"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries 

```{r}
library(plyr)
library(dplyr)
library(ggplot2)
library(mixtools)
library(fitdistrplus)
```

## Load data 

MRF = long-lived red sea urchin (Mesocentrotus franciscanus)  
LPI = short-lived green sea urchin (Lytechinus variegatus)


```{r}
MRF <- read.delim("../MRF_CpG.txt", header=FALSE) %>%
  dplyr::rename(Gene = V1) %>% dplyr::rename(CpG_ratio = V2)

LVA <- read.delim("../LVA_CpG.txt", header=FALSE) %>%
  dplyr::rename(Gene = V1) %>% dplyr::rename(CpG_ratio = V2)

SPU <- read.delim("../SPU_CpG.txt", header=FALSE) %>%
  dplyr::rename(Gene = V1) %>% dplyr::rename(CpG_ratio = V2)

LPICTUS <- read.delim("../LPICTUS_CpG.txt", header=FALSE) %>%
  dplyr::rename(Gene = V1) %>% dplyr::rename(CpG_ratio = V2)

set.seed(101)
```

## Choosing cut-offs 

```{r}
range(MRF$CpG_ratio) ## 0.00000 1.82588
hist(MRF$CpG_ratio)

MRF_filtered <- MRF %>% 
  filter(CpG_ratio >= 0.001) %>% # setting minimum
  filter(CpG_ratio <= 1.5) # setting maximum 

range(MRF_filtered$CpG_ratio) ## 0.00990199 1.49170000
hist(MRF_filtered$CpG_ratio) 
```


```{r}
range(LVA$CpG_ratio) ## 0.00000 2.41394
hist(LVA$CpG_ratio)

LVA_filtered <- LVA %>% 
  filter(CpG_ratio >= 0.001) %>% # setting minimum
  filter(CpG_ratio <= 1.5) # setting maximum 

range(LVA_filtered$CpG_ratio) ## 0.0249947 1.4963300
hist(LVA_filtered$CpG_ratio) 
```


```{r}
range(SPU$CpG_ratio) ## 0.0296791 3.2727300
hist(SPU$CpG_ratio)

SPU_filtered <- SPU %>% 
  filter(CpG_ratio >= 0.001) %>% # setting minimum
  filter(CpG_ratio <= 1.5) # setting maximum 

range(SPU_filtered$CpG_ratio) ## 0.0296791 1.4894400
hist(SPU_filtered$CpG_ratio) 
```


```{r}
range(LPICTUS$CpG_ratio) ## 0.00000 2.41394
hist(LPICTUS$CpG_ratio)

LPICTUS_filtered <- LPICTUS %>% 
  filter(CpG_ratio >= 0.001) %>% # setting minimum
  filter(CpG_ratio <= 1.5) # setting maximum 

range(LPICTUS_filtered$CpG_ratio) ## 0.0249947 1.4963300
hist(LPICTUS_filtered$CpG_ratio) 
```
Calculating number of rows for filtered df

```{r}
MRF_filtered ##22,183
LVA_filtered ##20,562
SPU_filtered ##20,875
LPICTUS_filtered ##28,534

MRF_filtered %>% write.csv("lab work/CpG_MRF_filtered.csv")
```



## Fitting mixture model with mixtools normalmixEM

https://dozenoaks.twelvetreeslab.co.uk/2019/06/mixture-models/

`normalmixEM` functions is telling it to find two gaussians in the observations

```{r}
mixmodel_MRF <- normalmixEM(MRF_filtered$CpG_ratio, k = 2)
summary(mixmodel_MRF) ## 481.04

mixmodel_LVA <- normalmixEM(LVA_filtered$CpG_ratio, k = 2)
summary(mixmodel_LVA) ## 318.14

mixmodel_SPU <- normalmixEM(SPU_filtered$CpG_ratio, k = 2)
summary(mixmodel_SPU) ## 561.98

mixmodel_LPICTUS <- normalmixEM(LPICTUS_filtered$CpG_ratio, k = 2)
summary(mixmodel_LPICTUS) ## 740.62

plot(mixmodel_MRF, which = 2, col2=c("darkred","darkgreen"), ylab1 = c(0, 1.5))
plot(mixmodel_LVA, which = 2, col2=c("darkgreen","darkred"), ylab1 = c(0, 1.5))
plot(mixmodel_SPU, which = 2, col2=c("darkred","darkgreen"), ylab2 = c(0, 1.5))
plot(mixmodel_LPICTUS, which = 2, col2=c("darkred","darkgreen"), ylab2 = c(0, 1.5))


```

### Plotting the above model

Defining function 

```{r}
fxn = function(mean, sd, lambda, n, binwidth) {
      stat_function(
        fun = function(x) {
          (dnorm(x, mean = mean, sd = sd)) * n * binwidth * lambda
        }
      )
    }
```


Plotting

```{r}
MRF_filtered %>% 
  ggplot(., aes(x=CpG_ratio)) + theme_classic() + 
  geom_histogram(binwidth=0.1, fill = "white", color="darkgrey", alpha=0.9) + ylim(0,5000) +
  ylab("Number of genes") + xlab("CpG Observed/Expected") +
  mapply(
    function(mean, sd, lambda, n, binwidth) {
      stat_function(
        fun = function(x) {
          (dnorm(x, mean = mean, sd = sd)) * n * binwidth * lambda
        }
      )
    },
    mean = mixmodel_MRF[["mu"]], #mean
    sd = mixmodel_MRF[["sigma"]], #standard deviation
    lambda = mixmodel_MRF[["lambda"]], #amplitude
    n = length(MRF_filtered$CpG_ratio), #sample size
    binwidth = 0.1 #binwidth used for histogram; needs to match the above binwidth 
  )

LVA_filtered %>% 
  ggplot(., aes(x=CpG_ratio)) + theme_classic() + 
  geom_histogram(binwidth=0.1, fill = "white", color="darkgrey", alpha=0.9) + ylim(0,5000) +
  ylab("Number of genes") + xlab("CpG Observed/Expected") +
  mapply(
    function(mean, sd, lambda, n, binwidth) {
      stat_function(
        fun = function(x) {
          (dnorm(x, mean = mean, sd = sd)) * n * binwidth * lambda
        }
      )
    },
    mean = mixmodel_LVA[["mu"]], #mean
    sd = mixmodel_LVA[["sigma"]], #standard deviation
    lambda = mixmodel_LVA[["lambda"]], #amplitude
    n = length(LVA_filtered$CpG_ratio), #sample size
    binwidth = 0.1 #binwidth used for histogram; needs to match the above binwidth 
  )

SPU_filtered %>% 
  ggplot(., aes(x=CpG_ratio)) + theme_classic() + 
  geom_histogram(binwidth=0.1, fill = "white", color="darkgrey", alpha=0.9) + ylim(0,5000) +
  ylab("Number of genes") + xlab("CpG Observed/Expected") +
  mapply(
    function(mean, sd, lambda, n, binwidth) {
      stat_function(
        fun = function(x) {
          (dnorm(x, mean = mean, sd = sd)) * n * binwidth * lambda
        }
      )
    },
    mean = mixmodel_SPU[["mu"]], #mean
    sd = mixmodel_SPU[["sigma"]], #standard deviation
    lambda = mixmodel_SPU[["lambda"]], #amplitude
    n = length(SPU_filtered$CpG_ratio), #sample size
    binwidth = 0.1 #binwidth used for histogram; needs to match the above binwidth 
  )

LPICTUS_filtered %>% 
  ggplot(., aes(x=CpG_ratio)) + theme_classic() + 
  geom_histogram(binwidth=0.1, fill = "white", color="darkgrey", alpha=0.9) + ylim(0,5000) +
  ylab("Number of genes") + xlab("CpG Observed/Expected") +
  mapply(
    function(mean, sd, lambda, n, binwidth) {
      stat_function(
        fun = function(x) {
          (dnorm(x, mean = mean, sd = sd)) * n * binwidth * lambda
        }
      )
    },
    mean = mixmodel_LPICTUS[["mu"]], #mean
    sd = mixmodel_LPICTUS[["sigma"]], #standard deviation
    lambda = mixmodel_LPICTUS[["lambda"]], #amplitude
    n = length(LPICTUS_filtered$CpG_ratio), #sample size
    binwidth = 0.1 #binwidth used for histogram; needs to match the above binwidth 
  )
```

## Finds intersection point of two component model

```{r}
intersect <- function(m1, s1, m2, s2, prop1, prop2){
  
  B <- (m1/s1^2 - m2/s2^2)
  A <- 0.5*(1/s2^2 - 1/s1^2)
  C <- 0.5*(m2^2/s2^2 - m1^2/s1^2) - log((s1/s2)*(prop2/prop1))
  
  (-B + c(1,-1)*sqrt(B^2 - 4*A*C))/(2*A)
}

mixmodel_MRF[["mu"]] #means
mixmodel_MRF[["sigma"]] #standard deviations
mixmodel_MRF[["lambda"]] #amplitudes

intersect_pt_MRF <- intersect(mixmodel_MRF[["mu"]][1], mixmodel_MRF[["sigma"]][1], 
                          mixmodel_MRF[["mu"]][2], mixmodel_MRF[["sigma"]][2], 
                          mixmodel_MRF[["lambda"]][1], mixmodel_MRF[["lambda"]][2])


mixmodel_LVA[["mu"]] #means
mixmodel_LVA[["sigma"]] #standard deviations
mixmodel_LVA[["lambda"]] #amplitudes

intersect_pt_LVA <- intersect(mixmodel_LVA[["mu"]][1], mixmodel_LVA[["sigma"]][1], 
                          mixmodel_LVA[["mu"]][2], mixmodel_LVA[["sigma"]][2], 
                          mixmodel_LVA[["lambda"]][1], mixmodel_LVA[["lambda"]][2])

mixmodel_SPU[["mu"]] #means
mixmodel_SPU[["sigma"]] #standard deviations
mixmodel_SPU[["lambda"]] #amplitudes

intersect_pt_SPU <- intersect(mixmodel_SPU[["mu"]][1], mixmodel_SPU[["sigma"]][1], 
                          mixmodel_SPU[["mu"]][2], mixmodel_SPU[["sigma"]][2], 
                          mixmodel_SPU[["lambda"]][1], mixmodel_SPU[["lambda"]][2])

mixmodel_LPICTUS[["mu"]] #means
mixmodel_LPICTUS[["sigma"]] #standard deviations
mixmodel_LPICTUS[["lambda"]] #amplitudes

intersect_pt_LPICTUS <- intersect(mixmodel_LPICTUS[["mu"]][1], mixmodel_LPICTUS[["sigma"]][1], 
                          mixmodel_LPICTUS[["mu"]][2], mixmodel_LPICTUS[["sigma"]][2], 
                          mixmodel_LPICTUS[["lambda"]][1], mixmodel_LPICTUS[["lambda"]][2])
```

## Tests fit of single component model

```{r}
fitdistr(MRF_filtered$CpG_ratio, "normal")
fitdistr(LVA_filtered$CpG_ratio, "normal")
fitdistr(SPU_filtered$CpG_ratio, "normal")
fitdistr(LPICTUS_filtered$CpG_ratio, "normal")
```

## Weakly and heavily methylated genes 

```{r}
MRF_filtered <- MRF_filtered %>%
  mutate(status = if_else(CpG_ratio >= intersect_pt_MRF[2], "weakly", "heavily"))

nrow(MRF_filtered) # 22,183
nrow(MRF_filtered[MRF_filtered$status == "weakly",]) # 12,891
nrow(MRF_filtered[MRF_filtered$status == "heavily",])# 9,292

LVA_filtered <- LVA_filtered %>%
  mutate(status = if_else(CpG_ratio >= intersect_pt_LVA[1], "weakly", "heavily"))

nrow(LVA_filtered) # 20,562
nrow(LVA_filtered[LVA_filtered$status == "weakly",]) # 13,045
nrow(LVA_filtered[LVA_filtered$status == "heavily",])# 7,517

SPU_filtered <- SPU_filtered %>%
  mutate(status = if_else(CpG_ratio >= intersect_pt_SPU[2], "weakly", "heavily"))

nrow(SPU_filtered) # 20,875
nrow(SPU_filtered[SPU_filtered$status == "weakly",]) # 14,218
nrow(SPU_filtered[SPU_filtered$status == "heavily",])# 6,657

LPICTUS_filtered <- LPICTUS_filtered %>%
  mutate(status = if_else(CpG_ratio >= intersect_pt_LPICTUS[2], "weakly", "heavily"))

nrow(LPICTUS_filtered) # 28,534
nrow(LPICTUS_filtered[LPICTUS_filtered$status == "weakly",]) # 16,590
nrow(LPICTUS_filtered[LPICTUS_filtered$status == "heavily",])# 11,944
```






